Modelling state-switching functional biologging data with hidden Markov models

### 2020-11-16 Report/Sections/intro.tex
address co-author edits thru 3.0

Biologging technology now provides researchers with data collected almost continuously in time \citep{Hooten:2017}.
The collection and analysis of data from devices such as accelerometers has brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}. In particular, the study of animal movement has been transformed by advanced tracking devices which collect biologging data \citep{Borger:2020}. Tags can record observations at over 50 times per second, resulting in time series containing millions of observations over the course of several hours. These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a new challenge for statisticians.

@@

Some FDA models account for between-curve dependence but are inadequate when modelling certain types of time dependencies. %, including that of Figure \ref{fig:data}.
For example, 
%multilevel models incorporate dependence via random effects. 
Previous studies have used multilevel models with random effects to capture variation between activity levels in children \citep{Morris:2007} and study differences in menstrual cycles \citet{Bromback:1998}. More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Chen:2012} and \citet{Crainiceanu:2009}. However, multilevel models are not appropriate in many biologging applications since they account for neither temporal dependence nor different curve types. 
%In the aforementioned study of children's activity levels \citep{Morris:2007}, the authors used a multilevel approach with a random effects to model variation between children. In a classic paper that introduced this multilevel structure,  \citet{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles. More recent work on multilevel models includes that of \citet{Di:2009}, \citet{Chen:2012} and \citet{Crainiceanu:2009}. However, multilevel models are not appropriate in many biologging applications since they account for neither temporal dependence nor different curve types. 
In addition to multilevel models, FDA researchers have also used functional time series to model dependence in sequences of curves. Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}. %While functional time series methods can in theory be applied to biologging data such as ours, 
However, functional time series do not account for sequences of time series curves that switch between a discrete number of well defined states.
%
Traditional FDA techniques similarly fall short when modelling complicated within-curve %acceleration 
data. In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function \citep{Yao:2005} 
%use this framework to construct a method of sparse longitudinal data analysis using functional principal component analysis. Alternatively, within-curve structure is sometimes generated 
or via random regression \citep{Rice:2001}. 
%For example, \citet{Rice:2001} use random regression to construct non-parametric mixed effects models for unequally sampled noisy curves. However, the combination of apparent behavioural changes combined with periodic behaviour makes accelerometer data such as ours difficult to model with these classic FDA techniques.
However, time series data exhibiting both sharp behavioural changes and periodic fine-scale structure is difficult to model with these classic FDA techniques.

@@

While useful, HMMs alone are also not sufficient to model intricate fine-scale time series data for three primary reasons.
%
Firstly, HMMs assume that subsequent observations are independent given an underlying hidden state process, but this is often not the case when observations are taken at extremely high frequencies. 
Several solutions have been proposed in the ecology literature such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. The CarHMM in particular explicitly models auto-correlation into an HMM while largely maintaining its base structure. 
%
Secondly, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales, (i.e., between and within curves). 
To deal with this issue animal movement modellers have employed hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}, which model processes at both scales with conditionally dependent HMMs.
%
Finally, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated dependence structures at short time scales. For example, \citet{Adam:2019} fail to capture fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}) using a traditional HHMM. \citet{Heerah:2017} use Fourier analysis within an HMM to account for %periodic behaviour in the context of describing 
daily behavioural cycles in marine mammals. In addition, Fourier analysis has previously been used with accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}. Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for fine-scale periodic structures.

%While animal movement modellers have taken steps to solve each of these issues independently, to our knowledge there is no unifying framework combining the three together. 
In this paper, we combine existing methods from statistical ecology literature in novel ways to account for complex temporally dependent functional data. The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models to model multi-scale functional and time-series data with intricate dependence structure.
%and thus extend the methods available for the analysis of functional data. 
We begin by describing HMMs as well as to two variants (CarHMMs and HHMMs) and discuss how Fourier analysis can handle fine-scale dependence structures. We also show how these methods can be combined to analyze increasingly complex data. In Section 3 we present a case study in which we fit several candidate models to data from a killer whale from the threatened northern resident population off the coast of British Columbia, Canada. Section 4 details a simulation study based on these candidate models, and in Section 5 we discuss our results.

>>>>

Biologging technology now provides researchers with kinematic data collected almost continuously in time \citep{Hooten:2017}.
The collection and analysis of data from devices such as accelerometers has brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}. In particular, the study of animal movement has been transformed by short-term, high-resolution suction-cup attached tags that record kinematic information about animal movement underwater \citep{Borger:2020,Dot:2016b}. Tags can record observations at over 50 times per second, resulting in time series containing millions of observations over the course of several hours. These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a new challenge for statisticians.

@@

Some FDA models account for between-curve dependence but are inadequate when modelling certain types of time dependencies. %, including that of Figure \ref{fig:data}.
For example, 
%multilevel models incorporate dependence via random effects. 
Previous studies have used multilevel models with random effects to capture variation between activity levels in children \citep{Morris:2007} and study differences in menstrual cycles \citet{Bromback:1998}. More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Chen:2012} and \citet{Crainiceanu:2009}. However, multilevel models are not appropriate in many biologging applications since they account for neither temporal dependence nor different curve types. 
%In the aforementioned study of children's activity levels \citep{Morris:2007}, the authors used a multilevel approach with a random effects to model variation between children. In a classic paper that introduced this multilevel structure,  \citet{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles. More recent work on multilevel models includes that of \citet{Di:2009}, \citet{Chen:2012} and \citet{Crainiceanu:2009}. However, multilevel models are not appropriate in many biologging applications since they account for neither temporal dependence nor different curve types. 
In addition to multilevel models, FDA researchers have also used functional time series to model dependence in sequences of curves. Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}. %While functional time series methods can in theory be applied to biologging data such as ours, 
However, functional time series do not account for sequences of time series curves that switch between a discrete number of well defined states.
%
Traditional FDA techniques similarly fall short when modelling complicated within-curve %acceleration 
data. In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function \citep{Yao:2005} 
%use this framework to construct a method of sparse longitudinal data analysis using functional principal component analysis. Alternatively, within-curve structure is sometimes generated 
or with random regression \citep{Rice:2001}. 
%For example, \citet{Rice:2001} use random regression to construct non-parametric mixed effects models for unequally sampled noisy curves. However, the combination of apparent behavioural changes combined with periodic behaviour makes accelerometer data such as ours difficult to model with these classic FDA techniques.
However, time series data exhibiting both sharp behavioural changes and periodic fine-scale structure is difficult to model with these classic FDA techniques.

@@

While useful, HMMs alone are also not sufficient to model intricate fine-scale time series data for three primary reasons.
%
Firstly, HMMs assume that subsequent observations are independent given an underlying hidden state process, but this is often not the case when observations are taken at extremely high frequencies. 
Several solutions have been proposed in the ecology literature such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. The CarHMM in particular explicitly models auto-correlation into an HMM while largely maintaining its base structure. 
%
Secondly, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales, (i.e., between and within curves). 
To address this issue, animal movement modellers have employed hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}, which model both scales with conditionally dependent HMMs.
%
Finally, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated dependence structures at short time scales. For example, \citet{Adam:2019} fail to capture fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}) using a traditional HHMM. \citet{Heerah:2017} use Fourier analysis within an HMM to account for %periodic behaviour in the context of describing 
daily behavioural cycles in marine mammals. In addition, Fourier analysis has previously been used with accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}. Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for fine-scale periodic structures.

%While animal movement modellers have taken steps to solve each of these issues independently, to our knowledge there is no unifying framework combining the three together. 
In this paper, we combine existing methods from statistical ecology literature in novel ways to account for complex, temporally dependent functional data. The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models to model multi-scale functional and time-series data with intricate dependence structure.
%and thus extend the methods available for the analysis of functional data. 
We begin by describing HMMs as well as to two variants (CarHMMs and HHMMs) and discuss how Fourier analysis can handle fine-scale dependence structures. We also show how these methods can be combined to analyze increasingly complex data. In Section 3 we present a case study in which we fit several candidate models to data from a killer whale from the threatened northern resident population off the coast of British Columbia, Canada. Section 4 details a simulation study based on these candidate models, and in Section 5 we discuss our results.

### 2020-11-16 Report/Sections/models.tex
address co-author edits thru 3.0

Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easy to calculate via the forward algorithm.
%
Let $y$ be the $T$-vector of the observed coarse-scale emissions and
$y^* \equiv \left\{y^*_1, \ldots,y^*_T\right\}$ be the collection of $T$ observed fine-scale emission vectors.
%
In addition, let $\theta^* \equiv \{\theta^{*(1)}, \ldots, \theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices. The likelihood of the observed data is then
%
\begin{equation}
    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*) \mathbf{1}_N
    \label{eqn:HHMM_likelihood}
\end{equation}
%
where $P(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_1)\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_1^*\right)$ for an HMM and $\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_1^*\right)$ for a CarHMM. The fine-scale likelihood $\calL_{\text{(Car)HMM}}$ corresponds to the desired fine-scale model. In addition, $P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to 
$f^{(i)}(y_t|y_{t-1})\calL_{\text{HMM}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_t^*\right)$ for both a CarHMM and an HMM.
%
For more information on specific considerations for HHMMs, such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see \citet{Adam:2019}.

@@

\subsection{General structure for building complex models}

@@

These examples include a few of many fine-scale models that can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models. For example, one may construct a three-level HHMM, where three scales of observations are recorded (fine-, medium-, and coarse-scale) and traditional HMMs are used to model all three levels. A myriad of possible models can be built using this framework, but these models quickly become complicated and computationally expensive to fit. Therefore, models should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.

>>>>

Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easily calculated using the forward algorithm.
%
Let $y$ be the $T$-vector of the observed coarse-scale emissions and
$y^* \equiv \left\{y^*_1, \ldots,y^*_T\right\}$ be the collection of $T$ observed fine-scale emission vectors.
%
In addition, let $\theta^* \equiv \{\theta^{*(1)}, \ldots, \theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices. The likelihood of the observed data is then
%
\begin{equation}
    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*) \mathbf{1}_N
    \label{eqn:HHMM_likelihood}
\end{equation}
%
where $P(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_1)\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_1^*\right)$ for an HMM and $\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_1^*\right)$ for a CarHMM. The fine-scale likelihood $\calL_{\text{(Car)HMM}}$ corresponds to the desired fine-scale model. In addition, $P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to 
$f^{(i)}(y_t|y_{t-1})\calL_{\text{HMM}}\left(\theta^{*(i)},
\Gamma^{*(i)};y_t^*\right)$ for both a CarHMM and an HMM.
%
For more information on specific considerations for HHMMs, such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see \citet{Adam:2019}.

@@

\subsection{Generalized hierarchical Markov models}

@@

These examples include a few of many fine-scale models that can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models. For example, one may construct a three-level HHMM, where three scales of observations are recorded (fine, medium, and coarse-scale) and dependent traditional HMMs are used to model all three levels. A myriad of possible models can be built using this framework, but these models quickly become complicated and computationally expensive to fit. Therefore, models should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.

### 2020-11-16 Report/Sections/results.tex
address co-author edits thru 3.0

To illustrate the process of constructing a model using these HMM building blocks, we investigate the dive behaviour of a northern resident killer whale (NRKW) off the coast of British Columbia, Canada, and construct several candidate models to describe the associated data.

Understanding animal behaviour is important for conservation efforts, as environmental changes often directly impact animal behaviour \citep{Sutherland:1998}. HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g. \citet{DeRuiter:2017} and \citet{Isojunno:2017}). For NRKWs, we are interested in categorizing different diving behaviours, with a particular focus on differentiating potential foraging dives. NRKWs prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKWs for predation \citep{Williams:2009,Noren:2011}. It is therefore of interest to determine the energetic expenditure of NRKWs foraging for Chinook as well as the frequency at which NRKWs catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}. Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that the behavioural state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is important to understand the true energetic requirements of a NRKW.

@@

The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz using Customizable Animal Solutions Tracking (CATs) tags (model number \#\#\#). The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left). Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. Tagging the killer whale caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored. We also ignore all dives between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm due to a malfunction in the tagging technology. To process the data, we smooth the accelerometer and depth data by taking a moving average within a window of $1/10^{th}$ of a second. We then define a killer whale ``dive" as any continuous interval of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}. Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, the vector of observed coarse-scale observations $y = \left\{y_1,\ldots,y_{267}\right\}$ is a time series of dive durations in seconds ($s$). For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \left\{y^*_{t,1},\ldots \right\}$, which is a vector of the within-dive acceleration data in meters per second squared ($m/s^2$). Naturally, the collection of all acceleration data is $y^* = \left\{y^*_1,\ldots,y^*_{267}\right\}$.

@@

The coarse-scale emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, travelling, and searching for prey.
On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as foraging and socialization \citep{Tennessen:2019b}. However, no dive in this data set has a maximum depth deeper than 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100 meters. It is therefore unlikely that any of the dives used in this study are successful foraging dives.

>>>>

To illustrate the process of constructing a model using these HMM building blocks, we investigate the dive behaviour of a northern resident killer whale off the coast of British Columbia, Canada, and construct several candidate models to categorize and describe its diving behaviour.

Understanding animal behaviour is important for conservation efforts, as environmental changes caused by anthropogenic activity can directly impact animal behaviour \citep{Sutherland:1998}. HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g. \citet{DeRuiter:2017} and \citet{Isojunno:2017}). For killer whales, we are interested in categorizing different diving behaviours to identify potential foraging dives. Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller prey \citep{Ford:2009}. Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook \citep{Williams:2009,Noren:2011,Wright:2017}. It is therefore of interest to determine the energetic expenditure of killer whales hunting for Chinook as well as the frequency at which they catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}. Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that the behavioural state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of a killer whale.

@@

The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm PDT and consist of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz using Customizable Animal Solutions Tracking (CATs) tags (model number \#\#\#). The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left). Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. Tagging the killer whale caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored. We also ignore all dives between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm due to a malfunction in the tagging technology. To process the data, we smooth the accelerometer and depth data by taking a moving average within a window of $1/10^{th}$ of a second. We then define a killer whale ``dive" as any continuous interval of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}. Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, the vector of observed coarse-scale observations $y = \left\{y_1,\ldots,y_{267}\right\}$ is a time series of dive durations in seconds ($s$). For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \left\{y^*_{t,1},\ldots \right\}$, which is a vector of the within-dive acceleration data in meters per second squared ($m/s^2$). Naturally, the collection of all acceleration data is $y^* = \left\{y^*_1,\ldots,y^*_{267}\right\}$.

@@

The coarse-scale emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}). 
Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, travelling, and searching for prey.
On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as hunting and socialization \citep{Tennessen:2019b}. However, no dive in this data set has a maximum depth deeper than 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100 meters. It is therefore unlikely that any of the dives used in this study are successful foraging dives.

### 2020-11-16 Report/main.tex
address co-author edits thru 3.0

% Title, authors, affiliations
%\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
\title[]{Modelling state-switching functional data with hidden Markov models}
\author{BlindedA\authorref{1}\thanksref{*}}
\author{BlindedB\authorref{2}}
\affiliation[1]{Author affiliations will go here in the accepted manuscript, 
but do NOT include them in your initial submission because it must be anonymous.}
\affiliation[2]{Second Affiliation}

@@

\iffalse
Advances in biologging technology have made a vast amount of high-frequency time series data available in a variety of fields. These data often exhibit complicated temporal dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA). Animal movement modellers often employ a hierarchical approach using hidden Markov models (HMMs) instead, but many intricate fine-scale processes violate basic assumptions of HMMs. We detail a general hierarchical approach based on HMMs which incorporates a wider class of models at fine scales. As a case study, we build a hierarchical model that combines an HMM with a moving-window auto-regressive process to describe the movement of a killer whale off the western coast of Canada. Our model produces more interpretable results as well as more accurate parameter and standard error estimates compared to existing baselines.
\fi

Sequences of curves sampled at high-frequency in time are increasingly common in practice, but they can exhibit complicated dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA). We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM). The distribution of each curve is then defined by another fine-scale model which may involve a variety of methods such as moving-window transformations, Fourier analysis, and auto-regression. As a case study, we use this framework to model the movement of a killer whale off the western coast of Canada. Through simulations we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing baselines. This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures.

>>>>

% Title, authors, affiliations
%\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
\title[]{Modelling state-switching functional biologging data with hidden Markov models}
\author{BlindedA\authorref{1}\thanksref{*}}
\author{BlindedB\authorref{2}}
\affiliation[1]{Author affiliations will go here in the accepted manuscript, 
but do NOT include them in your initial submission because it must be anonymous.}
\affiliation[2]{Second Affiliation}

@@

\iffalse
Advances in biologging technology have made a vast amount of high-frequency time series data available in a variety of fields. These data often exhibit complicated temporal dependence structures that cannot be modelled using common methods of Functional Data Analysis (FDA). Animal movement modellers often employ a hierarchical approach using hidden Markov models (HMMs) instead, but many intricate fine-scale processes violate basic assumptions of HMMs. We detail a general hierarchical approach based on HMMs which incorporates a wider class of models at fine scales. As a case study, we build a hierarchical model that combines an HMM with a moving-window auto-regressive process to describe the movement of a killer whale off the western coast of Canada. Our model produces more interpretable results as well as more accurate parameter and standard error estimates compared to existing baselines.
\fi

Data sets comprised of sequences of curves sampled at high-frequency in time are increasingly common in practice, but they can exhibit complicated dependence structures that cannot be modelled using common methods of Functional Data Analysis (FDA). We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM). The distribution of each curve is then defined by another fine-scale model which may involve a variety of methods such as moving-window transformations, Fourier analysis, and auto-regression. This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures. As a case study, we use this framework to model the fine-scale kinematic movement of a northern resident killer whale off the western coast of Canada. Through simulations we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing baselines.
